Local Spectral Clustering of Density Upper Level Sets

### 2019-04-29 arxiv/arxiv_local_density_clustering.tex
added arxiv version of paper

%%%%%%%% ICML 2019 submission %%%%%%%%%%%%%%%%%

\documentclass{article}

\usepackage{icml2019}
% \usepackage[accepted]{icml2019}

\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amsfonts, amsthm, amssymb}
\usepackage{graphicx}
\usepackage[parfill]{parskip}
\usepackage{enumerate}
\usepackage[shortlabels]{enumitem}
\usepackage{bm}

@@

\theoremstyle{aldenrmrk}
\newtheorem{remark}{Remark}
\newcommand{\theHalgorithm}{\arabic{algorithm}}

%%%%%
\icmltitlerunning{Local Spectral Clustering of Density Upper Level Sets}

@@

\twocolumn[
\icmltitle{Local Spectral Clustering of Density Upper Level Sets}
\icmlsetsymbol{equal}{*}

\begin{icmlauthorlist}
\icmlauthor{Alden Green}{cmu}
\icmlauthor{Sivaraman Balakrishnan}{cmu}
\icmlauthor{Ryan J. Tibshirani}{cmu}
\end{icmlauthorlist}

\icmlaffiliation{cmu}{Department of Statistics and Data Science, Carnegie Mellon
  University, Pittsburgh PA, USA} 
\icmlcorrespondingauthor{Alden Green}{ajgreen@andrew.cmu.edu}
\icmlkeywords{local clustering}

\vskip 0.3in
]

\printAffiliationsAndNotice{}

@@

When applied to geometric graphs constructed from a large number of samples,
global spectral clustering methods can be computationally cumbersome and   
insensitive to the local geometry of the underlying distribution
\citep{leskovec2010,mahoney2012}.  This has led to recent increased interest in
local spectral algorithms, which leverage locally-biased spectra computed using
random walks around a user-specified seed node.  A popular local clustering
algorithm is Personalized PageRank (PPR), first introduced by
\citet{haveliwala2003}, and further developed by
\citet{spielman2011,spielman2014,andersen2006,mahoney2012,zhu2013},
among others.

Local spectral clustering techinques have been practically very successful
\citep{leskovec2010,andersen2012,gleich2012,mahoney2012,wu2012}, which has led
many authors to develop supporting theory
\citep{spielman2013,andersen2009,gharan2012,zhu2013} that gives worst-case
guarantees on traditional graph-theoretic notions of cluster quality (like
conductance).  In this paper, we adopt a more traditional statistical viewpoint,
and examine what the output of a local clustering algorithm on $\Xbf$ reveals
about the unknown density $f$.  In particular, we examine the ability of the PPR
algorithm to recover \emph{density clusters} of $f$, which are defined as the
connected components of the upper level set $\{x \in \Rd : f(x) \geq \lambda\}$
for some threshold $\lambda > 0$ (a central object of central interest in the
classical statistical literature on clustering, dating back to
\citealt{hartigan1981}).

@@

Here we define a pair of criteria that reflect the quality of a cluster with
respect to $G=(V,E,w)$.  There are many graph-based measures of cluster quality  
that one could consider; see, e.g., \citet{yang2015,fortunato2010} for 
an overview.  The pair of criteria that we focus on are (arguably) quite
natural, and moreover, they play a fundamental role in our analysis 
of the PPR algorithm.  Our two criteria capture the \emph{external} and
\emph{internal} connectivity of a subset $S \subseteq V$, denoted $\Phi(S; G)$
and $\Psi(S; G)$, respectively, and defined below in turn.

@@

For $S \subseteq V$, denote by $G[S] = (S, E_S, w_S)$ the subgraph induced by 
$S$ (where the edges are $E_S = E \cap (S \times S)$). Let $\Abf_S,\Dbf_S$
be the adjacency matrix and degree matrix, respectively, of $G[S]$.  Define the
random walk matrix as usual, $\Wbf = \Dbf_S^{-1} \Abf_S$, and for $v \in V$,
write 
$$
q_{vu}^{(t)} = e_v\Wbf_S^t e_u
$$
for the $t$-step transition probability of a random walk over $G[S]$
originating at $v$.\footnote{Given a starting node $v$ and and a random walk
  defined by transition probability matrix $\mathbf{P}$, the rotation $e_v
  \mathbf{P}^t$ is used to denote the distribution of the random walk after $t$
  steps.}   Also write \smash{$\widetilde\pibf = (\widetilde\pi_{u})_{u \in S}$}
for the stationary distribution of this random walk.  (Given the definition of 
$\Wbf_S$, it is well-known that the stationary distribution is given by
\smash{$\widetilde\pi_{u} = (\Dbf_{S})_{uu} / \vol(S; G[S])$}.)

Our internal connectivity parameter will capture the time it takes for the
random walk over $G[S]$ to mix (approach the stationary distribution)
uniformly over $S$.  For this, we first define \emph{relative pointwise mixing
  time} of $G[S]$ as 
$$
\tau_{\infty}(G[S]) = \min\set{ m: \frac{|q_{vu}^{(m)} - 
      \widetilde\pi_u|}{\widetilde\pi_u} \leq \frac{1}{4}, 
    \; \text{for $u,v \in V$}},  
$$
%where \smash{$\qbf = (\qbf_v^{(1)}, \qbf_v^{(2)}, ...)_{v \in V}$}, and
%\smash{$\qbf_v^{(m)} = (q_{vu}^{(m)})_{u \in V}$}. 
% RJT: I got rid of the notational dependence on \qbf, because these quantities
% are in turn just defined in terms of the induced subgraph G[S]
% AJG: RESOLVED
Now our internal connectivity parameter is simply the inverse mixing time, 
\begin{equation}
\label{eqn: inv_mixing_time}
\Psi(S; G) = \frac{1}{\tau_{\infty}(G[S])}.
\end{equation}

If $S$ has normalized cut no greater than $\Phi$, and inverse mixing time no
less than $\Psi$, we call it as a \emph{$(\Phi,\Psi)$-cluster}. Both
local \citep{zhu2013} and global \citep{kannan04} spectral algorithms have been
shown to output clusters (or partitions) which approximate the optimal $(\Phi,
\Psi)$-cluster (or partition) for a given graph $G$.\footnote{In the case of
  \citet{kannan04}, the internal connectivity parameter $\phi$ is actually the
  conductance, i.e., the minimum normalized cut within the subgraph $G[S]$. See
  Theorem 3.1 in their paper for details; however, note that $\phi^2 /
  \log(\vol(S)) \leq O(\Psi)$, and so the lower bound on $\phi$ translates to a
  lower bound on $\Psi$.}

@@

For a level $\beta > 0$ and a target stationary measure $\pi_0 > 0$, we define a
\emph{$\beta$-sweep cut} of $\pbf$ as  
\begin{equation}
\label{eqn: sweep_cuts}
S_\beta = \{u \in V: p_u > \beta \pi_{0}\}.
\end{equation}
Having computed sweep cuts over a range \smash{$\beta \in  
(\frac{3}{10},\frac{1}{2})$},\footnote{The choice of a specific range such as 
\smash{$(\frac{3}{10}, \frac{1}{2})$} is standard in the analysis of PPR
algorithms, see, e.g., \citet{zhu2013}.}
we output a cluster \smash{$\widehat{C} = S_{\beta^*}$}, based on the sweep cut
$S_{\beta^*}$ that minimizes the normalized cut \smash{$\Phi(S_{\beta^*};
  G_{n,r})$} as defined in \eqref{eqn: norm_cut}. For concreteness, we summarize
this procedure in Algorithm \ref{alg: ppr}.

@@

\item Under largely the same set of geometric conditions, we derive in Theorem
  \ref{thm: inverse_mixing_time_lower_bound_nonconvex} a lower bound on the
  inverse mixing time of a random walk over $\Csig[\Xbf]$.  We also provide in 
  Theorem \ref{thm: inverse_mixing_time_lower_bound} a tighter lower bound, but
  under more restrictive assumptions (convexity of $\Csig$).

\item In Section \ref{sec: consistent_cluster_estimation_with_ppr}, we show in
  Theorem \ref{thm: consistent_recovery_of_density_clusters} that these bounds
  in bounds in Theorems \ref{thm: conductance_upper_bound} and \ref{thm:
    inverse_mixing_time_lower_bound_nonconvex},
  on the cluster quality criteria, have algorithmic consequences for PPR:
  properly initialized, Algorithm \ref{alg: ppr} performs consistent density
  cluster estimation in the sense of \eqref{eqn:
    consistent_density_cluster_recovery}.

\item We show in Corollary \ref{cor: ppr_cluster} that Theorems \ref{thm:
    conductance_upper_bound} and \ref{thm:
    inverse_mixing_time_lower_bound_nonconvex}, along with the results in 
  \citet{zhu2013}, lead to alternative, graph-theoretic guarantees on cluster 
  quality: an upper bound on the normalized cut of the estimated cluster 
  \smash{$\widehat{C}$}, and an upper bound on volume of the symmetric set
  difference between \smash{$\widehat{C}$} and $\Cset[\Xbf]$.

@@

Building on earlier work of
\citet{koltchinskii2000}, \citet{vonluxburg2008,hein2005} studied the limiting behaviour of spectral clustering
algorithms. These authors show that when samples are obtained from a
distribution, and we appropriately construct a geometric graph, the spectrum of
the Laplacian converges to that of the Laplace-Beltrami operator on the
data-manifold. However, relating the partition obtained using the
Laplace-Beltrami operator to the more intuitively defined high-density
clusters can be challenging in general.

Perhaps most similar to our results are the works
\citet{vempala2004,shi2009,schiebinger2015}, who study the consistency of
spectral algorithms in recovering the latent labels in certain parametric and
nonparametric mixture models. These results focus on global rather than local
algorithms, and as such impose global rather than local conditions on the nature
of the density. Moreover, they do not in general ensure recovery of density
clusters, which is the focus in our work.

@@

In order to provide meaningful bounds on the normalized cut and inverse mixing
time of an empirical density cluster, we must introduce conditions on the
density $f$. Let $B(x,r) = \{y \in \Rd: \norm{y - x} \leq r\}$ be the closed
ball of radius $r \geq 0$, centered at $x \in \Rd$.  Given a set $\Aset
\subseteq \Rd$ and $\sigma > 0$, define $\Asig = \Aset + B(0,\sigma) = \{y \in
\Rd: \inf_{x \in \Aset} \norm{y - x} \leq \sigma\}$, which we call the
$\sigma$-expansion of $\Aset$.

We are now ready to give the conditions, stated with respect to a density
cluster $\Cset \in \Cbb_f(\lambda)$, for some threshold $\lambda > 0$, and an
expansion parameter $\sigma > 0$.

@@

\item 
\label{asmp: low_noise_density}
\emph{Low noise density:} There exists $\gamma > 0$ such that for all $x
\in \Rd$ with $0 < \dist(x, \Csig) \leq \sigma$,   
$$
  \inf_{x' \in \Csig} f(x') - f(x) \geq  \dist(x, \Csig)^{\gamma},
$$
where \smash{$\dist(x,\Aset) = \inf_{x_0 \in \Aset} \norm{x - x_0}$}.
% RJT: should we change this to c_0?  c_1 seems to be used below and throughout
% for other constants, and c_0 is not used

\item
\label{asmp: cluster_separation}
\emph{Cluster separation:}
For all $\Cset' \in \Cbb_f(\lambda)$,
$$
\dist(\Csig,\Csig') > \sigma,
$$
where \smash{$\dist(\Aset,\Aset') = \inf_{x \in \Aset} \dist(x,\Aset')$}.

\item
\label{asmp: cluster_diameter}
\emph{Cluster diameter:}
There exists $D < \infty$ such that for all $x, x' \in \Csig$,
$$
\norm{x - x'} \leq D.
$$
\end{enumerate}

@@

Assumptions \ref{asmp: bounded_density}, \ref{asmp: low_noise_density}, and
\ref{asmp: cluster_separation} are used to upper bound $\Phi(\Cset[\Xbf];
G_{n,r})$, whereas \ref{asmp: bounded_density} and \ref{asmp: cluster_diameter} 
are required to lower bound $\Psi(\Cset[\Xbf]; G_{n,r})$. We note that the
lower bound on minimum density in \ref{asmp: bounded_density} along with \ref{asmp:  
cluster_separation} are similar to the $(\sigma,\epsilon)$-saliency of
\citet{chaudhuri2010}, a standard density clustering assumption, while
\ref{asmp: low_noise_density} is seen in, e.g., \citet{singh2009} (as well as
many other works on density clustering and level set estimation.) It is worth
highlighting that these assumptions are all local in nature, a benefit of
studying a local algorithm such as PPR.

In the next several subsections, we will derive bounds on the cluster quality
criteria evaluated on ($\sigma$-expansions of) density clusters. For notational simplicity,
hereafter for $S \subseteq V$, we will abbreviate $\Phi(S; G_{n,r})$ by
$\Phi_{n,r}(S)$, and similarly, $\Psi(S; G_{n,r})$ by $\Psi_{n,r}(S)$, and
$\tau_{\infty}(G_{n,r}[S])$ by $\tau_{n,r}(S)$. We will also use $\nu$ for
Lebesgue measure on $\Rd$, and $\nu_d = \nu(B)$ for the measure of the unit ball
$B=B(0,1)$.  
% RJT: you had B(0,d) before ... just check this should have been B(0,1) 
% AJG: Yep.

@@

We start with an upper bound on the normalized cut \eqref{eqn: norm_cut} of 
$\Cset_\sigma[\Xbf]$. (In the theorem, the upper bound on the density in
Assumption \ref{asmp: bounded_density} will not actually be needed, so we omit
the parameter $\Lambda_\sigma>0$ from the theorem statement.)

\begin{theorem}
\label{thm: conductance_upper_bound}
Fix $\lambda > 0$, and let $\Cset \in \Cbb_f(\lambda)$ satisfy
Assumptions \ref{asmp: bounded_density}, \ref{asmp: low_noise_density}, and
\ref{asmp: cluster_separation}, for some 
$\sigma, \lambda_{\sigma}, c_1, \gamma > 0$. Moreover, let
\begin{equation}
\label{eqn: weighted_cluster_volume}
\nu_{\Pbb}(\Csig) \leq \frac{1}{2}
\end{equation}
where $\nu_{\Pbb}(\Aset) = \int_{\Aset} f(x) dx$ for $\Aset \subset \Rd$.
 Then for any $0 < r <
\sigma/(4d)$, $0 < \delta < 1$, $\epsilon > 0$, and
\begin{equation}
\label{eqn: conductance_sample_complexity}
n \geq \frac{9\log(2/\delta)}{\epsilon^2}\left(\frac{1}
  { \lambda_{\sigma}^2\nu(\Csig) \nu_d r^d}\right)^2,
\end{equation}
we have
\begin{equation}
  \label{eqn: conductance_additive_error_bound}
  \frac{\Phi_{n,r}(\Csig[\mathbf{X}])}{r} \leq 4 c_{\sigma} d
  \frac{\lambda}{\lambda_{\sigma}} \frac{(\lambda_{\sigma} -
    \frac{r^{\gamma}}{\gamma+1})}{\lambda_{\sigma}} + \epsilon, 
\end{equation}
with probability at least $1-\delta$, where $c_{\sigma} = 1 / \sigma$. 
\end{theorem}

\begin{remark}
The proof of Theorem \ref{thm: conductance_upper_bound}, along with all other
proofs in this paper, can be found in the supplementary document. The key 
idea is that for any $x \in \Cset$, the simple (possibly loose) fact
$B(x,\sigma) \subseteq \Csig$ translates into the upper bound $\nu(\Cset_{\sigma
  + r}) \leq (1 + 2d r /\sigma)\nu(\Csig)$. We leverage \ref{asmp:
  low_noise_density} to find a corresponding bound on the weighted volume,
then apply standard concentration inequalities to convert from population-
to sample-based results.  
\end{remark}

@@

\begin{remark}
  The convexity condition is only needed to lower bound 
  \smash{$\widetilde{\Phi}_{\Pbb,r}$}. Any lower bound on
  \smash{$\widetilde{\Phi}_{\Pbb,r}$} could be directly plugged into the 
  machinery of the proof of Theorem \ref{thm: inverse_mixing_time_lower_bound}
  to yield an alternative result. In recent work, \citet{abbasi17} developed new
  population-level bounds on the conductance function (without requiring
  convexity), however the Markov chain dealt with there is somewhat different 
  than the one we consider.    
\end{remark}

@@

\section{Empirical Performance of PPR on Gaussian Mixture Models}
\label{sec: experiments}

The assumptions and theory of Section \ref{sec: cluster_quality_bounds} -- and therefore Section \ref{sec: consistent_cluster_estimation_with_ppr} as well-- are tailored towards density functions with sharp transitions in gradient around the perimeter of the density cluster. These type of functions will satisfy \ref{asmp: bounded_density} with a small ratio of $\Lambda_{\sigma}$ to $\lambda_{\sigma}$, while still satisfying \ref{asmp: low_noise_density}, potentially for $\gamma << 1$.

\begin{figure*}
	\begin{center}
		\centerline{\includegraphics[width=\columnwidth, height = 3 in]{graphic_1}}
		\caption{Algorithm \ref{alg: ppr} run over 4 example datasets. Parameters $\alpha$ and $\pi_0$ were tuned to minimize normalized cut of output set. Colors represent \pprspace cuts; shapes correspond to the maximum density among mixture components. Sample size $n = 400$.}
		\label{fig: graphic_1}
	\end{center}
\end{figure*}

As mentioned previously, one line of work on spectral algorithms assesses their performance on mixture models. Particularly well-developed is the characterization of clustering performance on Gaussian Mixture Models (GMMs). Gaussian Mixture Models have smooth derivatives, and are therefore not good candidates for our work (at least, for reasonable values of the parameters in \ref{asmp: bounded_density} - \ref{asmp: cluster_diameter} ). Since they are a classical choice in the clustering literature, we investigate \pprspace performance on them empirically.

Figure \ref{fig: graphic_1} shows data generated from four different Gaussian mixture models. The story follows roughly the same narrative as our theoretical work. The left two plots demonstrate instances, in one and two-dimensions, respectively, where the \pprspace cut has small error relative to a density cut. The right two plots, by contrast, show \pprspace failing to recover, even approximately, a density cut.

In the top right panel, the leftmost density cluster has a smaller cutsize than the set \pprspace ultimately recovers. However, the recovered set (colored in blue) has a higher volume, and therefore a lower normalized cut.

The bottom right panel shows the effect of a high diameter on the performance of \pprspace. Although any one vertex on the boundary of a density level set may have few cut edges, the length of the boundary is sufficiently long to make the entire cut too large for \pprspace to select it. Instead, the \pprspace cut defaults to a more geometrically compact shape, which has much smaller (unweighted) perimeter.

>>>>

%%%%%%%% Local Spectral Clustering of Density Upper Level Sets %%%%%%%%%%%%%%%%%

\documentclass{article}
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{booktabs}
\usepackage{xr-hyper}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amsfonts, amsthm, amssymb}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage[parfill]{parskip}
\usepackage{enumerate}
\usepackage[shortlabels]{enumitem}
\usepackage{bm}
\usepackage{fullpage}

\usepackage[square,sort,comma,numbers]{natbib}
\renewcommand{\bibname}{REFERENCES}
\renewcommand{\bibsection}{\subsubsection*{\bibname}}

@@

\theoremstyle{aldenrmrk}
\newtheorem{remark}{Remark}

@@

\title{Local Spectral Clustering of Density Upper Level Sets}
\author{Alden Green, Sivaraman Balakrishnan, and Ryan Tibshirani}
\date{\today}
\maketitle

@@

When applied to geometric graphs constructed from a large number of samples,
global spectral clustering methods can be computationally cumbersome and   
insensitive to the local geometry of the underlying distribution
\citep{leskovec2010,mahoney2012}.  This has led to recent increased interest in
local spectral algorithms, which leverage locally-biased spectra computed using
random walks around a user-specified seed node.  A popular local clustering
algorithm is Personalized PageRank (PPR), first introduced by
\citep{haveliwala2003}, and further developed by
\citep{spielman2011,spielman2014,andersen2006,mahoney2012,zhu2013},
among others.

Local spectral clustering techinques have been practically very successful
\citep{leskovec2010,andersen2012,gleich2012,mahoney2012,wu2012}, which has led
many authors to develop supporting theory
\citep{spielman2013,andersen2009,gharan2012,zhu2013} that gives worst-case
guarantees on traditional graph-theoretic notions of cluster quality (like
conductance).  In this paper, we adopt a more traditional statistical viewpoint,
and examine what the output of a local clustering algorithm on $\Xbf$ reveals
about the unknown density $f$.  In particular, we examine the ability of the PPR
algorithm to recover \emph{density clusters} of $f$, which are defined as the
connected components of the upper level set $\{x \in \Rd : f(x) \geq \lambda\}$
for some threshold $\lambda > 0$ (a central object of central interest in the
classical statistical literature on clustering, dating back to
\citet{hartigan1981}).

@@

Here we define a pair of criteria that reflect the quality of a cluster with
respect to $G=(V,E,w)$.  There are many graph-based measures of cluster quality  
that one could consider; see, e.g., \citep{yang2015,fortunato2010} for 
an overview.  The pair of criteria that we focus on are (arguably) quite
natural, and moreover, they play a fundamental role in our analysis 
of the PPR algorithm.  Our two criteria capture the \emph{external} and
\emph{internal} connectivity of a subset $S \subseteq V$, denoted $\Phi(S; G)$
and $\Psi(S; G)$, respectively, and defined below in turn.

@@

For $S \subseteq V$, denote by $G[S] = (S, E_S, w_S)$ the subgraph induced by 
$S$ (where the edges are $E_S = E \cap (S \times S)$). Let $\Abf_S,\Dbf_S$
be the adjacency matrix and degree matrix, respectively, of $G[S]$.  Define the
random walk matrix as usual, $\Wbf = \Dbf_S^{-1} \Abf_S$, and for $v \in V$,
write 
$$
q_{vu}^{(t)} = e_v\Wbf_S^t e_u
$$
for the $t$-step transition probability of a random walk over $G[S]$
originating at $v$.\footnote{Given a starting node $v$ and and a random walk
  defined by transition probability matrix $\mathbf{P}$, the notation $e_v
  \mathbf{P}^t$ is used to denote the distribution of the random walk after $t$
  steps.}   Also write \smash{$\widetilde\pibf = (\widetilde\pi_{u})_{u \in S}$}
for the stationary distribution of this random walk.  (Given the definition of 
$\Wbf_S$, it is well-known that the stationary distribution is given by
\smash{$\widetilde\pi_{u} = (\Dbf_{S})_{uu} / \vol(S; G[S])$}.)

Our internal connectivity parameter will capture the time it takes for the
random walk over $G[S]$ to mix (approach the stationary distribution)
uniformly over $S$.  For this, we first define the \emph{relative pointwise mixing
  time} of $G[S]$ as 
$$
\tau_{\infty}(G[S]) = \min\set{ m: \frac{|q_{vu}^{(m)} - 
      \widetilde\pi_u|}{\widetilde\pi_u} \leq \frac{1}{4}, 
    \; \text{for $u,v \in V$}}. 
$$
%where \smash{$\qbf = (\qbf_v^{(1)}, \qbf_v^{(2)}, ...)_{v \in V}$}, and
%\smash{$\qbf_v^{(m)} = (q_{vu}^{(m)})_{u \in V}$}. 
% RJT: I got rid of the notational dependence on \qbf, because these quantities
% are in turn just defined in terms of the induced subgraph G[S]
% AJG: RESOLVED
Now our internal connectivity parameter is simply the inverse mixing time, 
\begin{equation}
\label{eqn: inv_mixing_time}
\Psi(S; G) = \frac{1}{\tau_{\infty}(G[S])}.
\end{equation}

If $S$ has normalized cut no greater than $\Phi$, and inverse mixing time no
less than $\Psi$, we call it as a \emph{$(\Phi,\Psi)$-cluster}. Both
local \citep{zhu2013} and global \citep{kannan04} spectral algorithms have been
shown to output clusters (or partitions) which approximate the optimal $(\Phi,
\Psi)$-cluster (or partition) for a given graph $G$.\footnote{In the case of
  \citep{kannan04}, the internal connectivity parameter $\phi$ is actually the
  conductance, i.e., the minimum normalized cut within the subgraph $G[S]$. See
  Theorem 3.1 in their paper for details; however, note that $\phi^2 /
  \log(\vol(S)) \leq O(\Psi)$, and so the lower bound on $\phi$ translates to a
  lower bound on $\Psi$.}

@@

For a level $\beta > 0$ and a target stationary measure $\pi_0 > 0$, we define a
\emph{$\beta$-sweep cut} of $\pbf$ as  
\begin{equation}
\label{eqn: sweep_cuts}
S_\beta = \{u \in V: p_u > \beta \pi_{0}\}.
\end{equation}
Having computed sweep cuts over a range \smash{$\beta \in  
(\frac{3}{10},\frac{1}{2})$},\footnote{The choice of a specific range such as 
\smash{$(\frac{3}{10}, \frac{1}{2})$} is standard in the analysis of PPR
algorithms, see, e.g., \citep{zhu2013}.}
we output a cluster \smash{$\widehat{C} = S_{\beta^*}$}, based on the sweep cut
$S_{\beta^*}$ that minimizes the normalized cut \smash{$\Phi(S_{\beta^*};
  G_{n,r})$} as defined in \eqref{eqn: norm_cut}. For concreteness, we summarize
this procedure in Algorithm \ref{alg: ppr}.

@@

\item Under largely the same set of geometric conditions, we derive in Theorem
  \ref{thm: inverse_mixing_time_lower_bound_nonconvex} a lower bound on the
  inverse mixing time of a random walk over $\Csig[\Xbf]$.

\item In Section \ref{sec: consistent_cluster_estimation_with_ppr}, we show in
  Theorem \ref{thm: consistent_recovery_of_density_clusters} that these bounds
  in Theorems \ref{thm: conductance_upper_bound} and \ref{thm:
    inverse_mixing_time_lower_bound_nonconvex},
  on the cluster quality criteria, have algorithmic consequences for PPR:
  properly initialized, Algorithm \ref{alg: ppr} performs consistent density
  cluster estimation in the sense of \eqref{eqn:
    consistent_density_cluster_recovery}.

\item We show in Corollary \ref{cor: ppr_cluster} that Theorems \ref{thm:
    conductance_upper_bound} and \ref{thm:
    inverse_mixing_time_lower_bound_nonconvex}, along with the results in 
  \citep{zhu2013}, lead to alternative, graph-theoretic guarantees on cluster 
  quality: an upper bound on the normalized cut of the estimated cluster 
  \smash{$\widehat{C}$}, and an upper bound on volume of the symmetric set
  difference between \smash{$\widehat{C}$} and $\Cset[\Xbf]$.

@@

Building on earlier work of
\citep{koltchinskii2000}, \citep{vonluxburg2008,hein2005} studied the limiting behaviour of spectral clustering
algorithms. These authors show that when samples are obtained from a
distribution, and we appropriately construct a geometric graph, the spectrum of
the Laplacian converges to that of the Laplace-Beltrami operator on the
data-manifold. However, relating the partition obtained using the
Laplace-Beltrami operator to the more intuitively defined high-density
clusters can be challenging in general.

% AJG 4/29: I need to rewrite to explain how its similar, because
% this is a good opportunity to show that implications
% of the type given in our work are broadly of interest.

Perhaps most similar to our results are the works
\citep{vempala2004,shi2009,schiebinger2015}, who study the consistency of
spectral algorithms in recovering the latent labels in certain parametric and
nonparametric mixture models. These results focus on global rather than local
algorithms, and as such impose global rather than local conditions on the nature
of the density. Moreover, they do not in general ensure recovery of density
clusters, which is the focus in our work.

@@

In order to provide meaningful bounds on the normalized cut and inverse mixing
time of an empirical density cluster, we must introduce conditions on the
density $f$. Let $B(x,r) = \{y \in \Rd: \norm{y - x} \leq r\}$ be the closed
ball of radius $r > 0$, centered at $x \in \Rd$.  Given a set $\Aset
\subseteq \Rd$ and $\sigma > 0$, define $\Asig = \Aset + B(0,\sigma) = \{y \in
\Rd: \inf_{x \in \Aset} \norm{y - x} \leq \sigma\}$, which we call the
$\sigma$-expansion of $\Aset$. For a differentiable function $g: \Rd \to \Rd$, write $\nabla g(x)$ to denote the Jacobian of $g$ evaluated at $x \in \Rd$.

We are now ready to give our required conditions, stated with respect to a density
cluster $\Cset \in \Cbb_f(\lambda)$ for some threshold $\lambda > 0$, and an
expansion parameter $\sigma > 0$.

@@

\item 
\label{asmp: low_noise_density}
\emph{Low noise density:} There exists $\gamma,c_0 > 0$ such that for all $x
\in \Rd$ with $0 < \dist(x, \Csig) \leq \sigma$,   
$$
  \inf_{x' \in \Csig} f(x') - f(x) \geq  c_0 \dist(x, \Csig)^{\gamma},
$$
where \smash{$\dist(x,\Aset) = \inf_{x_0 \in \Aset} \norm{x - x_0}$}.

\item
\label{asmp: cluster_separation}
\emph{Cluster separation:}
For all $\Cset' \in \Cbb_f(\lambda)$ with $\Cset' \not= \Cset$,
$$
\dist(\Csig,\Csig') > \sigma,
$$
where \smash{$\dist(\Aset,\Aset') = \inf_{x \in \Aset} \dist(x,\Aset')$}.

\item
\label{asmp: cluster_diameter}
\emph{Cluster diameter:}
There exists $D < \infty$ such that for all $x, x' \in \Csig$,
$$
\norm{x - x'} \leq D.
$$

\item
\label{asmp: biLipschitz_mapping}
\emph{biLipschitz mapping to convex set:}
There exists $\mathcal{K} \subseteq \Rd$ convex, and $g: \Reals^d \to \Reals^d$ satisfying
\begin{equation*}
\det(\nabla g (x) ) = 1,\frac{1}{L}\norm{x - y} \leq \norm{g(x) - g(y)} \leq L \norm{x - y} ~ \text{for all $x,y \in \Reals^d$}
\end{equation*}
such that $\Csig$ is the image of $\mathcal{K}$ by $g$, $\Csig = g(\mathcal{K})$.
\end{enumerate}

@@

Assumptions \ref{asmp: bounded_density}, \ref{asmp: low_noise_density},
\ref{asmp: cluster_separation} and \ref{asmp: biLipschitz_mapping} are used to upper bound $\Phi(\Cset[\Xbf];
G_{n,r})$, whereas \ref{asmp: bounded_density}, \ref{asmp: cluster_diameter} and \ref{asmp: biLipschitz_mapping} are required to lower bound $\Psi(\Cset[\Xbf]; G_{n,r})$. We note that the
lower bound on minimum density in \ref{asmp: bounded_density} along with \ref{asmp:  
cluster_separation} are similar to the $(\sigma,\epsilon)$-saliency of
\citep{chaudhuri2010}, a standard density clustering assumption, while
\ref{asmp: low_noise_density} is seen in, e.g., \citep{singh2009} (as well as
many other works on density clustering and level set estimation.)  It is worth
highlighting that these assumptions are all local in nature, a benefit of
studying a local algorithm such as \ppr.

%%% AJG 4/29: My attempt to explain why our results aren't fatuous.

We also note that while many of these geometric conditions are typical in the density clustering literature, the restrictions we will impose upon them in order to obtain meaningful implications for \pprspace will not be. This is natural. The spectral algorithm we consider is not specifically designed for the task of level set estimation, and in fact one should expect \pprspace to fail to recover -- either in the sense of \eqref{eqn: consistent_density_cluster_recovery}, or indeed any reasonable notion of cluster recovery -- a density cluster of sufficiently large diameter or sufficiently small thickness (though we do not provide any lower bounds to this effect). Indeed, one of the primary motivations of this work was to better understand and characterize the distinctions between those level sets which are well conditioned for spectral algorithms, and those which are not.

In the next several subsections, we will derive bounds on the cluster quality
criteria evaluated on ($\sigma$-expansions of) density clusters. For notational simplicity,
hereafter for $S \subseteq V$, we will abbreviate $\Phi(S; G_{n,r})$ by
$\Phi_{n,r}(S)$, and similarly, $\Psi(S; G_{n,r})$ by $\Psi_{n,r}(S)$, and
$\tau_{\infty}(G_{n,r}[S])$ by $\tau_{n,r}(S)$. We will also use $\nu$ for
Lebesgue measure on $\Rd$, and $\nu_d = \nu(B)$ for the measure of the unit ball
$B=B(0,1)$.

@@

We start with an upper bound on the normalized cut \eqref{eqn: norm_cut} of 
$\Cset_\sigma[\Xbf]$. (In Theorem \ref{thm: conductance_upper_bound}, the upper bound on the density in
Assumption \ref{asmp: bounded_density} will not actually be needed, so we omit
the parameter $\Lambda_\sigma>0$ from the theorem statement.) For $\mathcal{S} \subseteq \Reals^d$ and $r > 0$, let
\begin{equation*}
\pi_{\Pbb,r}(\mathcal{S}) := \frac{\int_{\mathcal{S}} \Pbb(B(x,r)) f(x) dx }{\int_{\Rd} \Pbb(B(x,r)) f(x) dx}.
\end{equation*}

\begin{theorem}
\label{thm: conductance_upper_bound}
Fix $\lambda > 0$, and let $\Cset \in \Cbb_f(\lambda)$ satisfy
Assumptions \ref{asmp: bounded_density}, \ref{asmp: low_noise_density},
and \ref{asmp: cluster_separation}, for some 
$\sigma, \lambda_{\sigma}, c_0, \gamma > 0$. Let $0 < r \leq \sigma/4d$ be such that
\begin{equation}
\label{eqn: weighted_cluster_volume}
\pi_{\Pbb,r}(\Csig) \leq \frac{1}{2}.
\end{equation}
 Then for any $0 < \delta < 1$, $\epsilon > 0$, if
\begin{equation}
\label{eqn: conductance_sample_complexity}
n \geq \frac{(2+\epsilon)^2\log(3/\delta)}{\epsilon^2}\left(\frac{25}
  {6 \lambda_{\sigma}^2\nu(\Csig) \nu_d r^d}\right)^2,
\end{equation}
then
\begin{equation}
  \label{eqn: conductance_additive_error_bound}
  \frac{\Phi_{n,r}(\Csig[\mathbf{X}])}{r} \leq \frac{4d}{\sigma}
  \frac{\lambda}{\lambda_{\sigma}} \frac{(\lambda_{\sigma} -
    c_0\frac{r^{\gamma}}{\gamma+1})}{\lambda_{\sigma}} + \epsilon, 
\end{equation}
with probability at least $1-\delta$.
\end{theorem}

\begin{remark}
The proof of Theorem \ref{thm: conductance_upper_bound}, along with all other
proofs in this paper, can be found in the supplementary document. The key 
idea is that for any $x \in \Cset$, the simple (possibly loose) fact
$B(x,\sigma) \subseteq \Csig$ translates into the upper bound $\nu(\Csig + rB) \leq (1 + 2d r /\sigma)\nu(\Csig)$. We leverage \ref{asmp:
  low_noise_density} to find a corresponding bound on the weighted volume,
then apply standard concentration inequalities to convert from population-
to sample-based results.  
\end{remark}

@@

\begin{remark}
  The convexity condition is only needed to lower bound 
  \smash{$\widetilde{\Phi}_{\Pbb,r}$}. Any lower bound on
  \smash{$\widetilde{\Phi}_{\Pbb,r}$} could be directly plugged into the 
  machinery of the proof of Theorem \ref{thm: inverse_mixing_time_lower_bound}
  to yield an alternative result. In recent work, \citep{abbasi-yadkori2016} developed new
  population-level bounds on the conductance function (without requiring
  convexity), however the Markov chain dealt with there is somewhat different 
  than the one we consider.    
\end{remark}

@@

\section{Experiments}
\label{sec: experiments}
